---
title: "Differential Transcript Expression & Differential Gene Expression"
author: "Emily F. Watts"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r install-packages, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
BiocManager::install("edgeR", update = FALSE, ask = FALSE)
```

```{r load-libraries}
library(edgeR)
library(ggplot2)
library(readr)

setwd("/Volumes/sheynkman/projects/LRP_Mohi_project")
```
Load count data - Transcripts
```{r read-data}
counts <- read.csv("/Volumes/sheynkman/projects/Mohi_MDS_LRP/01_isoseq/collapse/merged.collapsed.flnc_count.txt", 
                   header = TRUE, row.names = 1)

group <- factor(c("Q157R", "Q157R", "Q157R", "WT", "WT", "WT"))

dge_raw <- DGEList(counts = counts, group = group)
```

Visualize raw counts
```{r raw-visualization}
boxplot(dge_raw$counts, main = "Boxplot of Raw Counts", las = 2, col = as.numeric(group))
plotMDS(dge_raw, col = as.numeric(group), main = "MDS Plot for Raw Counts")
plotMD(dge_raw, col = as.numeric(group), main = "MD Plot for Raw Counts")
abline(h = 0, col = "red", lty = 2, lwd = 2)
```

MD (mean-difference) plots per sample
```{r md-plots}
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (i in 1:ncol(dge_raw$counts)) {
  plotMD(dge_raw, column = i, main = paste("MD Plot for Sample", i))
  abline(h = 0, col = "red", lty = 2, lwd = 2)
}
par(mfrow = c(1, 1)) # Reset plotting area to single panel
```

Filter lowly expressed transcripts (required for modeling)
```{r filter}
keep <- filterByExpr(dge_raw)
table(keep)

dge <- dge_raw[keep, , keep.lib.sizes = FALSE]
```

Normalize Counts with TMM
```{r normalize}
dge <- calcNormFactors(dge, method = "TMM")
```

Visualize normalized counts
```{r normalized-visualization}
boxplot(dge$counts, main = "Boxplot of Normalized Counts", las = 2, col = as.numeric(group))
plotMDS(dge, col = as.numeric(group), main = "MDS Plot for Normalized Counts")
plotMD(dge, col = as.numeric(group), main = "MD Plot for Normalized Counts")
abline(h = 0, col = "red", lty = 2, lwd = 2)
```

Design matrix and dispersion
```{r design-dispersion}
design <- model.matrix(~ group)
rownames(design) <- colnames(dge)

dge_disp <- estimateDisp(dge, design)
dge_disp$common.dispersion

plotBCV(dge_disp)
```

Fit model and test
```{r fit-test}
fit <- glmQLFit(dge_disp, design)
plotQLDisp(fit)

result <- glmQLFTest(fit, coef = 2)
topTags(result)
```

FDR and DEG Summary
```{r fdr-deg-summary}
FDR <- p.adjust(result$table$PValue, method = "BH")
sum(FDR < 0.05)

qlf <- glmQLFTest(fit)
top <- rownames(topTags(qlf))
cpm(dge_disp)[top, ]

summary(decideTests(qlf))
plotMD(qlf)
abline(h = c(-1, 1), col = "blue")
```

Save DEG results & intermediate files
```{r save-deg-results}
deg_results <- topTags(result, n = Inf)$table
write.table(deg_results, 
            file = "/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/transcript_DEG_results.txt", 
            sep = "\t", quote = FALSE, row.names = TRUE)

# Raw and normalized CPM Matrices
write.table(cpm(dge_raw), file = "/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/raw_CPM_matrix.txt", sep = "\t", quote = FALSE)
write.table(cpm(dge), file = "/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/normalized_CPM_matrix.txt", sep = "\t", quote = FALSE)

# List of filtered-in transcripts
write.table(rownames(dge), file = "/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/filtered_transcripts.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

# Design matrix to keep record of the model used 
write.table(design, file = "/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/design_matrix.txt", sep = "\t", quote = FALSE, row.names = TRUE, col.names = NA)

# Top genes table 
top_genes <- topTags(qlf, n = 50)$table
write.table(top_genes, file = "/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/top_transcripts.txt", sep = "\t", quote = FALSE, row.names = TRUE)

# Save dipersion plots (BCV and QL)
png("/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/BCV_plot.png", width = 800, height = 600)
plotBCV(dge_disp)
dev.off()

png("/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/QLDisp_plot.png", width = 800, height = 600)
plotQLDisp(fit)
dev.off()

# MDS plot (normalized)
png("/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/MDS_plot.png", width = 800, height = 600)
plotMDS(dge, col = as.numeric(group), main = "MDS Plot for Normalized Counts")
dev.off()
```

Volcano Plot - left = downregulated in Q157R vs. WT (hihger in WT); right = up-regulated in Q157R vs. WT (higher in Q157R)
```{r volcano-plot}
deg_results$Gene <- rownames(deg_results)

ggplot(deg_results, aes(x = logFC, y = -log10(PValue))) + 
  geom_point(aes(color = FDR < 0.05)) + 
  scale_color_manual(values = c("black", "red")) + 
  theme_minimal() + 
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 P-value")

ggsave("/Volumes/sheynkman/projects/LRP_Mohi_project/19_LRP_summary/edgeR/transcript_volcano_plot.png", width = 6, height = 6, dpi = 300)
```


Interactive volcano plot
```{r interactive-volcano-plot, message=FALSE, warning=FALSE}
library(ggiraph)
library(plotly)

gg <- ggplot(deg_results, aes(x = logFC, y = -log10(PValue),
                              tooltip = Gene, data_id = Gene)) +
  geom_point_interactive(aes(color = FDR < 0.05)) +
  scale_color_manual(values = c("black", "red")) +
  theme_minimal() +
  labs(title = "Interactive Volcano Plot", x = "Log2 Fold Change", y = "-Log10 P-value")

girafe(ggobj = gg, width_svg = 10, height_svg = 6)
```
















