---
title: "20_DTE_DGE"
author: "Emily F. Watts, adapted from Madison Mehlferber"
date: "2025-04-29"
output: html_document
---
  
```{r setup, include=FALSE}
# read in the needed packages 
library(edgeR)
library(here)
library(splines)
library(tidyverse)
library(rtracklayer)
library(knitr)
library(pheatmap)
library(RColorBrewer)
```
Define input and output directories and set up needed new directories
```{r}
# Helper function to create directories if missing
create_dir_if_needed <- function(path) {
  if (!dir.exists(path)) {
    dir.create(path, recursive = TRUE)
    cat("Created directory:", path, "\n")
  } else {
    cat("Directory already exists:", path, "\n")
  }
}
# Define base project path
base_project_dir <- "/Volumes/sheynkman/projects/Mohi_MDS_LRP"

# Define sub paths
counts_file <- file.path(base_project_dir, "01_isoseq", "collapse", "merged.collapsed.flnc_count.txt")
sqanti_classification_table <- file.path(base_project_dir, "03_filter_sqanti", "filtered_MDS_classification.txt")
gene_map_file <- file.path(base_project_dir, "01_reference_tables", "ensg_gene.tsv")
project_root <- file.path(base_project_dir, "20_DTE_DTU")

# Define output folders
results_folder <- file.path(project_root, "results")
experiment_subfolder <- file.path(results_folder, "differential_transcript_expression")
plots_subfolder <- file.path(experiment_subfolder, "plots")
output_table_subfolder <- file.path(experiment_subfolder, "output_table")

# Create folders
create_dir_if_needed(results_folder)
create_dir_if_needed(experiment_subfolder)
create_dir_if_needed(plots_subfolder)
create_dir_if_needed(output_table_subfolder)

# Read in data
counts <- read.csv(counts_file, header=TRUE, row.names=1)
sqanti_table <- read.delim(sqanti_classification_table, header = TRUE, sep = "\t")
gene_map <- read.table(gene_map_file, header = TRUE, sep = "\t", col.names = c("associated_gene", "gene_name"))
```
Differential Expression Analysis: Q157R vs WT. 
In a most analyses, a stringent cutoff (e.g., <0.05 or <0.1) is recommended.
Using FDR < 0.1 here.
calcNormFactors(object,
                method = c("TMM","TMMwsp","RLE","upperquartile","none"),
                refColumn = NULL, logratioTrim = .3, sumTrim = 0.05, doWeighting = TRUE,
                Acutoff = -1e10, p = 0.75, ...)
```{r}
# Define the sample groups (must match the columns in the counts matrix)
group <- factor(c("Q157R", "Q157R", "Q157R", "WT", "WT", "WT"))

# Create DGEList object
dge_transcript <- DGEList(counts = counts, group = group)

# Filter lowly expressed transcripts
dge_transcript <- dge_transcript[filterByExpr(dge_transcript), , keep.lib.sizes = FALSE]

# Normalize library sizes
dge_transcript <- calcNormFactors(dge_transcript)

# Design matrix for group comparison
design <- model.matrix(~ group)

# Estimate dispersion
dge_transcript <- estimateDisp(dge_transcript, design)

# Fit the model and perform the quasi-likelihood F-test
fit <- glmQLFit(dge_transcript, design)
result <- glmQLFTest(fit, coef = 2)

# Extract all differential expression results
deg_transcripts <- topTags(result, n = Inf)$table

# Save full DE results
write.table(deg_transcripts, 
            file = file.path(output_table_subfolder, "transcript_DEG_results.txt"),
            sep = "\t", quote = FALSE, row.names = TRUE)

# Extract top 500 transcripts
top_500_tab <- deg_transcripts %>% 
  slice_min(order_by = FDR, n = 500)

# Check significance by p-value and FDR
tab_pval <- top_500_tab %>% filter(PValue < 0.05)
tab_fdr  <- top_500_tab %>% filter(FDR < 0.1)

# Add isoform IDs as a new column (if not already)
library(tibble)  # tidyverse already loads it, but just noting
top_500_tab <- rownames_to_column(top_500_tab, var = "isoform")

# Merge with SQANTI3 to get associated gene names
tab_merged <- left_join(top_500_tab, sqanti_table %>% select(isoform, associated_gene), by = "isoform")

# Save merged table
write.table(tab_merged, 
            file = file.path(output_table_subfolder, "transcripts_top500_merged.txt"), 
            sep = "\t", row.names = FALSE, quote = FALSE)

# How many unique genes are represented?
unique_genes <- unique(tab_merged$associated_gene)
cat("Number of unique genes among top 500 transcripts:", length(unique_genes), "\n")

# Display the first 100 genes in a neat table
top_100_genes <- data.frame(Gene_Name = head(unique_genes, 100))
knitr::kable(top_100_genes)

# Display all unique genes in a 5-column table
gene_matrix <- matrix(unique_genes, ncol = 5, byrow = TRUE)
gene_matrix_df <- as.data.frame(gene_matrix)
colnames(gene_matrix_df) <- c("Gene_Name_1", "Gene_Name_2", "Gene_Name_3", "Gene_Name_4", "Gene_Name_5")
knitr::kable(gene_matrix_df)

# Find non-unique genes (possible isoform switching)
non_unique_genes <- tab_merged %>% 
  filter(duplicated(associated_gene) | duplicated(associated_gene, fromLast = TRUE)) %>%
  arrange(associated_gene)

# Save non-unique gene entries
write.table(non_unique_genes, 
            file = file.path(output_table_subfolder, "transcripts_top500_nonunique.txt"), 
            sep = "\t", row.names = FALSE, quote = FALSE)
```
Save differential expression results
```{r}
# Save all DE transcripts
write.table(deg_transcripts, 
            file = file.path(output_table_subfolder, "transcript_DEG_results_all.txt"), 
            sep = "\t", quote = FALSE, row.names = TRUE)

# Filter DE transcripts based on FDR < 0.1
deg_transcripts_filtered <- deg_transcripts %>% filter(FDR < 0.1)

# Save filtered DE transcripts
write.table(deg_transcripts_filtered, 
            file = file.path(output_table_subfolder, "transcripts_DEG_results_filtered_FDR01.txt"), 
            sep = "\t", quote = FALSE, row.names = TRUE)

# Print number of filtered transcripts
cat("Number of DE transcripts with FDR < 0.1:", nrow(deg_transcripts_filtered), "\n")
```

```{r}
# ------------------------------------------
# Analyze top DE transcripts
# ------------------------------------------

# Extract top 500 DE transcripts (from deg_transcripts already sorted by FDR)
top_500_tab <- deg_transcripts %>%
  slice_min(order_by = FDR, n = 500) %>%
  rownames_to_column(var = "isoform")

# Filter top 500 for low p-values or FDR
top_500_pval <- top_500_tab %>% filter(PValue < 0.01)
top_500_fdr  <- top_500_tab %>% filter(FDR < 0.1)

# Merge with SQANTI to get gene and transcript names
tab_merged <- left_join(top_500_tab, 
                        sqanti_table %>% select(isoform, associated_gene, associated_transcript), 
                        by = "isoform")

# Identify unique genes
unique_gene_names <- unique(tab_merged$associated_gene)
cat("Number of unique genes in top 500 transcripts:", length(unique_gene_names), "\n")

# ------------------------------------------
# Display Top Genes
# ------------------------------------------

# 1. Single-column table (first 100)
top_100_gene_df <- data.frame(Gene_Name = head(unique_gene_names, 100))
knitr::kable(top_100_gene_df)

# 2. Multi-column table (5 columns)
if (length(unique_gene_names) >= 5) {
  num_rows <- ceiling(length(unique_gene_names) / 5)
  gene_names_matrix <- matrix(c(unique_gene_names, rep(NA, num_rows * 5 - length(unique_gene_names))), 
                              ncol = 5, byrow = TRUE)
  gene_table_df <- as.data.frame(gene_names_matrix)
  colnames(gene_table_df) <- paste0("Gene_Name_", 1:5)
  knitr::kable(gene_table_df)
} else {
  cat("Fewer than 5 unique genes found — skipping 5-column format.\n")
}
```

Finding the non-unique values in a df in r - postulated that these may be candidates of isoform switching events & write output tables
```{r}
# ------------------------------------------
# Find genes with multiple DE isoforms (potential isoform switching)
# ------------------------------------------

# Identify genes appearing more than once in the top 500 DE isoforms
multi_isoform_genes <- tab_merged %>%
  filter(duplicated(associated_gene) | duplicated(associated_gene, fromLast = TRUE)) %>%
  arrange(associated_gene)

# Print summary
cat("Number of transcripts with multi-isoform genes:", nrow(multi_isoform_genes), "\n")
cat("Number of unique genes with multiple isoforms DE:",
    length(unique(multi_isoform_genes$associated_gene)), "\n")

# ------------------------------------------
# Save merged DE table and multi-isoform subset
# ------------------------------------------

# Save full top 500 DE table merged with SQANTI
write.table(tab_merged, 
            file = file.path(output_table_subfolder, "transcripts_DE_top500_merged_with_genes.txt"), 
            sep = "\t", quote = FALSE, row.names = FALSE)

# Save multi-isoform subset
write.table(multi_isoform_genes, 
            file = file.path(output_table_subfolder, "transcripts_DE_multi_isoform_genes_top500.txt"), 
            sep = "\t", quote = FALSE, row.names = FALSE)
```
Add gene names, load, and reshape expression data
```{r}
# ------------------------------------------
# Load dynamic transcripts (top 500 DE)
# ------------------------------------------
dynamic_transcripts <- read.table(
  file.path(output_table_subfolder, "transcripts_DE_top500_merged_with_genes.txt"), 
  header = TRUE, sep = "\t"
)

# ------------------------------------------
# Add gene names and display names (using already loaded tables)
# ------------------------------------------
# Add display_name to sqanti_table
sqanti_table <- sqanti_table %>%
  mutate(display_name = case_when(
    !is.na(associated_transcript) & associated_transcript != "novel" ~ associated_transcript,
    grepl("^novelGene_", associated_gene) ~ associated_gene,
    TRUE ~ associated_gene
  ))

# Merge gene map and display name
dynamic_transcripts <- dynamic_transcripts %>%
  left_join(gene_map, by = "associated_gene") %>%
  left_join(sqanti_table[, c("isoform", "display_name")], by = "isoform")

# Clean up version suffixes
dynamic_transcripts$display_name <- gsub("\\.\\d+$", "", dynamic_transcripts$display_name)
gene_map$gene_name <- gsub("\\.\\d+$", "", gene_map$gene_name)

# ------------------------------------------
# Generate CPM matrix and reshape
# ------------------------------------------
dge <- DGEList(counts = counts, group = group)
cpm_matrix <- cpm(dge, log = FALSE)

cpm_df <- as.data.frame(cpm_matrix) %>%
  rownames_to_column("transcript_id")

cpm_long <- cpm_df %>%
  pivot_longer(cols = -transcript_id, names_to = "sample", values_to = "CPM") %>%
  mutate(group = case_when(
    str_detect(sample, "BioSample_[1-3]") ~ "Q157R",
    str_detect(sample, "BioSample_[4-6]") ~ "WT",
    TRUE ~ NA_character_
  ))

# ------------------------------------------
# Merge with dynamic transcripts metadata
# ------------------------------------------
dynamic_cpm_long <- cpm_long %>%
  filter(transcript_id %in% dynamic_transcripts$isoform) %>%
  left_join(dynamic_transcripts %>% 
              select(isoform, associated_gene, gene_name, display_name),
            by = c("transcript_id" = "isoform"))
```
Plot isoform expression for a gene of interest (Q157R vs WT)
```{r}
# Choose gene of interest (update as needed)
unique(dynamic_cpm_long$associated_gene)[1:10] # Can use this to get a gene of interest to test
gene_of_interest <- "ENSMUSG00000013736.17"  # Replace with your gene ID

# Look up gene symbol (if available)
symbol <- gene_map %>%
  filter(associated_gene == gene_of_interest) %>%
  pull(gene_name) %>%
  unique()

# Build plot title
plot_title <- if (length(symbol) > 0 && !is.na(symbol) && symbol != "") {
  paste("Isoform Expression for", symbol, "(WT vs Q157R)")
} else {
  paste("Isoform Expression for", gene_of_interest, "(WT vs Q157R)")
}

# Extract data for plotting
gene_plot_data <- dynamic_cpm_long %>%
  filter(associated_gene == gene_of_interest)

# Use default group ordering (Q157R on the left)
gene_plot_data$group <- factor(gene_plot_data$group)  # keep existing order

# Plot isoform-level expression
ggplot(gene_plot_data, aes(x = group, y = CPM, color = group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.7) +
  facet_wrap(~ display_name, scales = "free_y") +
  theme_minimal() +
  labs(title = plot_title,
       x = "Condition",
       y = "CPM") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(size = 10),
        plot.title = element_text(face = "bold"))
```
Plot logCPM for genes with multiple DE isoforms
```{r}
# Define group labels from logFC direction
dynamic_transcripts$group <- ifelse(dynamic_transcripts$logFC > 0, "Q157R", "WT")
dynamic_transcripts$group <- factor(dynamic_transcripts$group, levels = c("WT", "Q157R"))

# Identify genes with >1 DE isoform
multi_isoform_genes <- dynamic_transcripts %>%
  filter(duplicated(associated_gene) | duplicated(associated_gene, fromLast = TRUE)) %>%
  arrange(associated_gene)

genes_to_plot <- unique(multi_isoform_genes$associated_gene)

# Loop through each gene and plot
for (gene in genes_to_plot) {
  gene_data <- dynamic_transcripts %>% filter(associated_gene == gene)

  # Skip if logCPM is missing
  if (!"logCPM" %in% colnames(gene_data) || all(is.na(gene_data$logCPM))) next

  # Gene symbol fallback
  gene_symbol <- unique(gene_data$gene_name)[1]

  # Build plot title
  title_label <- if (!is.na(gene_symbol) && gene_symbol != "") {
    paste("Expression of", gene_symbol, "Isoforms (WT vs Q157R)")
  } else {
    paste("Expression of", gene, "Isoforms (WT vs Q157R)")
  }

  # Build plot
  p <- ggplot(gene_data, aes(x = group, y = logCPM, color = group)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.7) +
    facet_wrap(~ display_name, scales = "free_y") +
    theme_minimal() +
    labs(title = title_label,
         x = "Condition", 
         y = "logCPM") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text = element_text(size = 10),
          plot.title = element_text(face = "bold"))

  # Clean filename (gene symbol or fallback)
  safe_name <- gsub("[:/\\\\\\s]", "_", ifelse(!is.na(gene_symbol), gene_symbol, gene))
  filename <- paste0(safe_name, "_isoform_expression_WT_vs_Q157R.pdf")

  # Save plot
  ggsave(file.path(plots_subfolder, filename),
         plot = p, width = 6, height = 5)
}
p
```
Create Heatmap
```{r}
# ------------------------------------------
# Prepare for Heatmaps
# ------------------------------------------

# Define annotation color mapping once
annotation_colors <- list(
  Group = c("Q157R" = "#336B87", "WT" = "#E99787")
)

# Define group annotations based on sample names
sample_names <- unique(dynamic_cpm_long$sample)
annotation_col <- data.frame(
  Group = ifelse(str_detect(sample_names, "BioSample_[1-3]"), "Q157R", "WT")
)
rownames(annotation_col) <- sample_names

# ------------------------------------------
# Heatmap for Top 500 Dynamic Transcripts
# ------------------------------------------

top_500_transcripts <- dynamic_transcripts %>%
  arrange(PValue) %>%
  slice_head(n = 500)

transcript_matrix_top500 <- dynamic_cpm_long %>%
  filter(transcript_id %in% top_500_transcripts$isoform) %>%
  select(transcript_id, sample, CPM) %>%
  pivot_wider(names_from = sample, values_from = CPM) %>%
  column_to_rownames("transcript_id")

transcript_matrix_log2_top500 <- log2(transcript_matrix_top500 + 1)

p_transcripts_top500 <- pheatmap(transcript_matrix_log2_top500,
                                 cluster_rows = FALSE,
                                 cluster_cols = TRUE,
                                 scale = "row",
                                 show_rownames = FALSE,
                                 annotation_col = annotation_col,
                                 annotation_colors = annotation_colors,
                                 main = "Top 500 Dynamic Transcripts (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_transcript_heatmap_top500_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p_transcripts_top500)
dev.off()

# ------------------------------------------
# Heatmap for Top 500 Dynamic Genes
# ------------------------------------------

top_500_genes <- top_500_transcripts %>%
  filter(!is.na(associated_gene) & associated_gene != "") %>%
  distinct(associated_gene, .keep_all = TRUE) %>%
  pull(associated_gene)

gene_matrix_top500 <- dynamic_cpm_long %>%
  filter(associated_gene %in% top_500_genes) %>%
  group_by(associated_gene, sample) %>%
  summarise(mean_CPM = mean(CPM, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sample, values_from = mean_CPM) %>%
  filter(!is.na(associated_gene) & associated_gene != "") %>%
  distinct(associated_gene, .keep_all = TRUE) %>%
  column_to_rownames("associated_gene")

gene_matrix_log2_top500 <- log2(gene_matrix_top500 + 1)

p_genes_top500 <- pheatmap(gene_matrix_log2_top500,
                           cluster_rows = FALSE,
                           cluster_cols = TRUE,
                           scale = "row",
                           show_rownames = TRUE,
                           annotation_col = annotation_col,
                           annotation_colors = annotation_colors,
                           main = "Top 500 Dynamic Genes (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_gene_heatmap_top500_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p_genes_top500)
dev.off()

# ------------------------------------------
# Heatmap for Top 50 Dynamic Transcripts
# ------------------------------------------

top_50_transcripts <- dynamic_transcripts %>%
  arrange(PValue) %>%
  slice_head(n = 50)

transcript_matrix_top50 <- dynamic_cpm_long %>%
  filter(transcript_id %in% top_50_transcripts$isoform) %>%
  select(transcript_id, sample, CPM) %>%
  pivot_wider(names_from = sample, values_from = CPM) %>%
  column_to_rownames("transcript_id")

transcript_matrix_log2_top50 <- log2(transcript_matrix_top50 + 1)

p_transcripts_top50 <- pheatmap(transcript_matrix_log2_top50,
                                cluster_rows = TRUE,
                                cluster_cols = TRUE,
                                scale = "row",
                                show_rownames = TRUE,
                                annotation_col = annotation_col,
                                annotation_colors = annotation_colors,
                                main = "Top 50 Dynamic Transcripts (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_transcript_heatmap_top50_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p_transcripts_top50)
dev.off()

# ------------------------------------------
# Heatmap for Top 50 Dynamic Genes
# ------------------------------------------

top_50_genes <- top_50_transcripts %>%
  filter(!is.na(associated_gene) & associated_gene != "") %>%
  distinct(associated_gene, .keep_all = TRUE) %>%
  pull(associated_gene)

gene_matrix_top50 <- dynamic_cpm_long %>%
  filter(associated_gene %in% top_50_genes) %>%
  group_by(associated_gene, sample) %>%
  summarise(mean_CPM = mean(CPM, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sample, values_from = mean_CPM) %>%
  filter(!is.na(associated_gene) & associated_gene != "") %>%
  distinct(associated_gene, .keep_all = TRUE) %>%
  column_to_rownames("associated_gene")

gene_matrix_log2_top50 <- log2(gene_matrix_top50 + 1)

p_genes_top50 <- pheatmap(gene_matrix_log2_top50,
                          cluster_rows = TRUE,
                          cluster_cols = TRUE,
                          scale = "row",
                          show_rownames = TRUE,
                          annotation_col = annotation_col,
                          annotation_colors = annotation_colors,
                          main = "Top 50 Dynamic Genes (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_gene_heatmap_top50_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p_genes_top50)
dev.off()
```
Calculate fractional abundance per transcript and identify top 50 most differential isoforms
```{r}
# ------------------------------------------
# Calculate average fractional abundance per gene
# ------------------------------------------

frac_long <- dynamic_cpm_long %>%
  filter(!is.na(associated_gene)) %>%
  group_by(associated_gene, sample) %>%
  mutate(gene_cpm = sum(CPM, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(fractional_abundance = CPM / gene_cpm)

# Average across replicates by condition
library(rlang)

average_frac_abundance <- frac_long %>%
  group_by(associated_gene, transcript_id, group) %>%
  summarise(avg_abundance = mean(fractional_abundance, na.rm = TRUE), .groups = "drop")

# Rename columns for clarity and compatibility
colnames(average_frac_abundance)[colnames(average_frac_abundance) == "associated_gene"] <- "Gene"
colnames(average_frac_abundance)[colnames(average_frac_abundance) == "transcript_id"] <- "Transcript"

# ------------------------------------------
# Filter to multi-isoform genes before plotting
# ------------------------------------------
genes_with_multiple_transcripts <- average_frac_abundance %>%
  group_by(Gene) %>%
  summarise(num_isoforms = n_distinct(Transcript), .groups = "drop") %>%
  filter(num_isoforms >= 2) %>%
  pull(Gene)

average_frac_abundance_filtered <- average_frac_abundance %>%
  filter(Gene %in% genes_with_multiple_transcripts)

# ------------------------------------------
# Pivot and prepare Top 50 Isoforms
# ------------------------------------------
wide_abundance <- average_frac_abundance_filtered %>%
  pivot_wider(
    names_from = group,
    values_from = avg_abundance,
    values_fill = list(avg_abundance = 0)
  )

# Force coercion to numeric in case of Rle/list types
wide_abundance <- wide_abundance %>%
  mutate(
    Q157R = as.numeric(unlist(Q157R)),
    WT = as.numeric(unlist(WT)),
    diff = abs(Q157R - WT)
  )

str(wide_abundance)  

top50_isoforms <- as.data.frame(wide_abundance) %>%
  arrange(desc(diff)) %>%
  head(50)

# Convert back to long format for plotting
top50_long <- top50_isoforms %>%
  pivot_longer(cols = c(Q157R, WT), names_to = "group", values_to = "avg_abundance") %>%
  mutate(isoform_label = paste(Gene, Transcript, sep = ": "))

# ------------------------------------------
# Plot
# ------------------------------------------
p_top50_isoforms <- ggplot(top50_long, aes(x = reorder(isoform_label, -avg_abundance), y = avg_abundance, fill = group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 50 Most Differentially Expressed Isoforms (WT vs Q157R)",
       x = "Isoform (Gene:Transcript)", y = "Average Fractional Abundance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_y_continuous(expand = c(0, 0))

p_top50_isoforms

ggsave(file.path(plots_subfolder, "Top50_Differential_Isoforms_WT_vs_Q157R.pdf"),
       p_top50_isoforms, width = 12, height = 8)

```
Create stacked bar plots
```{r}
# ------------------------------------------
# Stacked Bar Plots for Isoform Usage per Gene
# ------------------------------------------
# Define group colors (not fill colors)
custom_group_colors <- c("Q157R" = "#336B87", "WT" = "#E99787")

# Get genes to plot
genes_to_plot_filtered <- unique(average_frac_abundance_filtered$Gene)

for (gene in genes_to_plot_filtered) {
  gene_data <- average_frac_abundance_filtered %>% filter(Gene == gene)

  if (nrow(gene_data) == 0) next

  # Get gene symbol if available
  gene_symbol <- gene_map %>%
    filter(associated_gene == gene) %>%
    pull(gene_name) %>%
    unique() %>%
    .[1]

  plot_title <- if (!is.na(gene_symbol) && gene_symbol != "") {
    paste("Stacked Isoform Usage for", gene_symbol, "(WT vs Q157R)")
  } else {
    paste("Stacked Isoform Usage for", gene, "(WT vs Q157R)")
  }

  # Consistent fill colors per gene (with color palette scaling by isoform count)
  isoform_palette <- colorRampPalette(brewer.pal(8, "Set2"))(length(unique(gene_data$Transcript)))

  # Plot
  p <- ggplot(gene_data, aes(x = group, y = avg_abundance, fill = Transcript)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = plot_title,
         x = "Group", y = "Average Fractional Abundance") +
    theme_minimal() +
    scale_fill_manual(values = isoform_palette) +
    guides(fill = guide_legend(reverse = TRUE)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          plot.title = element_text(face = "bold")) +
    scale_y_continuous(expand = c(0, 0))

  # Safe file name
  safe_name <- gsub("[:/\\\\\\s]", "_", ifelse(!is.na(gene_symbol), gene_symbol, gene))

  # Save
  ggsave(file.path(plots_subfolder, paste0(safe_name, "_stacked_barplot_WT_vs_Q157R.pdf")),
         p, width = 6, height = 4)
}
p
```
Gene level plotting 
```{r}
# 2. Filter the averaged fractional abundance to these genes
average_frac_abundance_filtered <- average_frac_abundance %>%
  filter(Gene %in% genes_to_plot)

# 3. Define green and brown colors
custom_colors <- c("Q157R" = "#336B87", "WT" = "#E99787")

# 4. Loop through genes and plot stacked bar plots
for (gene in genes_to_plot) {
  gene_data <- average_frac_abundance_filtered %>% filter(Gene == gene)
  
  if (nrow(gene_data) == 0) next
  
  p <- ggplot(gene_data, aes(x = group, y = avg_abundance, fill = Transcript)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = paste("Stacked Barplot for", gene, "(WT vs Q157R)"),
         x = "Group", y = "Average Fractional Abundance") +
    theme_minimal() +
    guides(fill = guide_legend(reverse = TRUE)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values = colorRampPalette(c("#336B87", "#E99787"))(length(unique(gene_data$Transcript))))  
  
  # Save each plot
  ggsave(file.path(plots_subfolder, paste0(gene, "_stacked_barplot_Q157R_vs_WT.pdf")),
         plot = p, width = 6, height = 5)
}
p
```
Replicate level plotting
```{r}
# 1. Loop through genes again, now for replicate-level plotting
for (gene in genes_to_plot) {
  gene_data_rep <- frac_long %>%
    filter(associated_gene == gene) %>%
    mutate(Sample_Group = paste(group, sample, sep = "_"))
  
  if (nrow(gene_data_rep) == 0) next
  
  # 2. Define colors across isoforms again
  n_transcripts <- length(unique(gene_data_rep$transcript_id))
  mycolors_rep <- colorRampPalette(c("#336B87", "#E99787"))(n_transcripts)  
  
  # 3. Plot
  p_rep <- ggplot(gene_data_rep, aes(x = Sample_Group, y = fractional_abundance, fill = transcript_id)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = paste("Replicate-level Stacked Barplot for", gene, "(Q157R vs WT)"),
         x = "Sample", 
         y = "Fractional Abundance") +
    theme_minimal() +
    scale_fill_manual(values = mycolors_rep) +
    guides(fill = guide_legend(reverse = TRUE, title = "Transcript")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(expand = c(0, 0))
  
  # Save replicate plot
  ggsave(file.path(plots_subfolder, paste0(gene, "_replicate_stacked_barplot_WT_vs_Q157R.pdf")),
         plot = p_rep, width = 8, height = 5)
}
p_rep
```