---
title: 'Heatmaps'
author: "Emily F. Watts"
date: "`r Sys.Date()`"
output: html_document
---

## Load packages
```{r load-packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(pheatmap)
  library(here)
  library(ggiraph)
  library(ggrepel)
})
```

#############################################################
####### Differential Expression (DTE & DGE) Heatmaps ########
#############################################################

## Load Transcript CPM & DGE data
```{r load-inputs, include=FALSE}
tx_df <- read_tsv("transcript_DEG_summary_table.tsv") %>%
  mutate(row_id = make.unique(Index))

cpm_cols <- c("A258_Q157R_CPM", "X504_Q157R_CPM", "A309_Q157R_CPM",
              "V335_WT_CPM", "V334_WT_CPM", "A310_WT_CPM")

cpm_matrix <- tx_df %>%
  select(row_id, all_of(cpm_cols)) %>%
  column_to_rownames("row_id") %>%
  as.matrix()

cpm_matrix <- cpm_matrix[, cpm_cols]  # Enforce sample order

# Modified transcript labeling: extract PB accession and mark novel transcripts
tx_label_map <- tx_df %>%
  mutate(
    # Try to extract PB accession from Transcript field
    pb_from_transcript = str_extract(Transcript, "PB\\.[0-9]+\\.[0-9]+"),
    # Try to extract PB accession from Index field (common alternative)
    pb_from_index = str_extract(Index, "PB\\.[0-9]+\\.[0-9]+"),
    # Use whichever one has the PB accession, fallback to Index if neither
    pb_accession = coalesce(pb_from_transcript, pb_from_index, Index),
    # Mark as novel if it doesn't contain ENSMUS
    is_novel = !str_detect(Transcript, "ENSMUS"),
    # Create final label: PB accession for all, + asterisk for novel only
    final_label = ifelse(is_novel, paste0(pb_accession, "*"), pb_accession)
  ) %>%
  select(row_id, final_label) %>%
  deframe()
```

## Load gene CPM & DGE data
```{r load-gene-data, include=FALSE}
gene_df <- read_tsv("gene_DEG_summary_table.tsv") %>%
  mutate(row_id = make.unique(Index))

gene_cpm <- gene_df %>%
  select(row_id, all_of(cpm_cols)) %>%
  column_to_rownames("row_id") %>%
  as.matrix()

gene_cpm <- gene_cpm[, cpm_cols]  # Enforce sample order

gene_label_map <- gene_df %>%
  select(row_id, Gene) %>%
  deframe()
```

## Define Subsets
```{r subset-top-features}
top_tx_df <- tx_df %>%
  mutate(is_sig = FDR_DEG < 0.05, abs_logFC = abs(logFC_DEG)) %>%
  arrange(desc(is_sig), desc(abs_logFC))

top100_tx <- top_tx_df %>% slice_head(n = 100) %>% pull(row_id)
top50_tx <- head(top100_tx, 50)

top_gene_df <- gene_df %>%
  mutate(is_sig = FDR_DEG < 0.05, abs_logFC = abs(logFC)) %>%
  arrange(desc(is_sig), desc(abs_logFC))

top100_genes <- top_gene_df %>% slice_head(n = 100) %>% pull(row_id)
top50_genes <- head(top100_genes, 50)
```

## Heatmap function for transcript CPM
```{r plot-tx-heatmap}
plot_tx_heatmap <- function(tx_ids, title) {
  mat <- cpm_matrix[rownames(cpm_matrix) %in% tx_ids, , drop = FALSE]
  mat <- log2(mat + 1)
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  # Use the modified label map with PB accessions and asterisks
  rownames(mat) <- tx_label_map[rownames(mat)]

  annotation_df <- data.frame(
    sample_name = colnames(mat),
    group = ifelse(grepl("WT", colnames(mat)), "WT", "Q157R")
  ) %>% column_to_rownames("sample_name")

  annotation_colors <- list(group = c(Q157R = "#336B87", WT = "#E99787"))

  col_order <- c(grep("Q157R", colnames(mat), value = TRUE),
                 grep("WT", colnames(mat), value = TRUE))
  mat <- mat[, col_order]
  annotation_df <- annotation_df[col_order, , drop = FALSE]

  n_rows <- nrow(mat)
  plot_height <- min(10, 5 + n_rows * 0.15)
  fontsize_row <- if (n_rows <= 30) 10 else if (n_rows <= 60) 8 else 6

  pheatmap(mat, scale = "row", cluster_rows = TRUE, cluster_cols = FALSE,
           annotation_col = annotation_df, annotation_colors = annotation_colors,
           show_colnames = TRUE, show_rownames = TRUE, main = title,
           fontsize_row = fontsize_row, fontsize_col = 10, height = plot_height)
}
```

## Heatmap function for gene CPM
```{r plot-gene-heatmap}
plot_gene_heatmap <- function(gene_ids, title) {
  mat <- gene_cpm[rownames(gene_cpm) %in% gene_ids, , drop = FALSE]
  mat <- log2(mat + 1)
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  rownames(mat) <- gene_label_map[rownames(mat)]

  annotation_df <- data.frame(
    sample_name = colnames(mat),
    group = ifelse(grepl("WT", colnames(mat)), "WT", "Q157R")
  ) %>% column_to_rownames("sample_name")

  annotation_colors <- list(group = c(Q157R = "#336B87", WT = "#E99787"))

  col_order <- c(grep("Q157R", colnames(mat), value = TRUE),
                 grep("WT", colnames(mat), value = TRUE))
  mat <- mat[, col_order]
  annotation_df <- annotation_df[col_order, , drop = FALSE]

  n_rows <- nrow(mat)
  plot_height <- min(10, 5 + n_rows * 0.15)
  fontsize_row <- if (n_rows <= 30) 10 else if (n_rows <= 60) 8 else 6

  pheatmap(mat, scale = "row", cluster_rows = TRUE, cluster_cols = FALSE,
           annotation_col = annotation_df, annotation_colors = annotation_colors,
           show_colnames = TRUE, show_rownames = TRUE, main = title,
           fontsize_row = fontsize_row, fontsize_col = 10, height = plot_height)
}
```

## Plot Top Transcript & Gene Heatmaps
```{r plot-top-heatmaps}
plot_tx_heatmap(top50_tx, "Top 50 Differentially Expressed Transcripts")
plot_tx_heatmap(top100_tx, "Top 100 Differentially Expressed Transcripts")
plot_gene_heatmap(top50_genes, "Top 50 Differentially Expressed Genes")
plot_gene_heatmap(top100_genes, "Top 100 Differentially Expressed Genes")
```

#############################################################
####### Differential Transcript Usage (DTU) Heatmaps ########
#############################################################
## Load DTU Data
```{r load-dtu, include=FALSE}
tx_dtu_df <- read_tsv("transcript_DTU_summary.tsv") %>%
  mutate(row_id = make.unique(Index))

frac_cols <- c("A258_Q157R_Frac", "X504_Q157R_Frac", "A309_Q157R_Frac",
               "V335_WT_Frac", "V334_WT_Frac", "A310_WT_Frac")

frac_matrix <- tx_dtu_df %>%
  select(row_id, all_of(frac_cols)) %>%
  column_to_rownames("row_id") %>%
  as.matrix()

frac_matrix <- frac_matrix[, frac_cols]

# Modified transcript labeling: extract PB accession and mark novel transcripts
frac_label_map <- tx_dtu_df %>%
  mutate(
    # Try to extract PB accession from Transcript field
    pb_from_transcript = str_extract(Transcript, "PB\\.[0-9]+\\.[0-9]+"),
    # Try to extract PB accession from Index field (common alternative)
    pb_from_index = str_extract(Index, "PB\\.[0-9]+\\.[0-9]+"),
    # Use whichever one has the PB accession, fallback to Index if neither
    pb_accession = coalesce(pb_from_transcript, pb_from_index, Index),
    # Mark as novel if it doesn't contain ENSMUS
    is_novel = !str_detect(Transcript, "ENSMUS"),
    # Create final label: PB accession for all, + asterisk for novel only
    final_label = ifelse(is_novel, paste0(pb_accession, "*"), pb_accession)
  ) %>%
  select(row_id, final_label) %>%
  deframe()

# Summarize DTU to gene level
gene_dtu_df <- tx_dtu_df %>%
  group_by(Gene) %>%
  summarise(
    row_id = first(Gene),
    lf_DTU = sum(lf_DTU, na.rm = TRUE),
    adj_p.value_DTU = min(adj.p.value_DTU, na.rm = TRUE),
    across(all_of(frac_cols), mean, na.rm = TRUE),
    .groups = "drop"
  )

gene_frac_matrix <- gene_dtu_df %>%
  select(row_id, all_of(frac_cols)) %>%
  column_to_rownames("row_id") %>%
  as.matrix()

gene_frac_matrix <- gene_frac_matrix[, frac_cols]

gene_frac_label_map <- gene_dtu_df %>%
  select(row_id, Gene) %>%
  deframe()
```

## Define DTU Subsets
```{r subset-dtu}
top_tx_dtu_df <- tx_dtu_df %>%
  arrange(adj.p.value_DTU, desc(abs(lf_DTU)))

top100_tx_dtu <- top_tx_dtu_df %>% slice_head(n = 100) %>% pull(row_id)
top50_tx_dtu <- head(top100_tx_dtu, 50)

top_gene_dtu_df <- gene_dtu_df %>%
  arrange(adj_p.value_DTU, desc(abs(lf_DTU)))

top100_gene_dtu <- top_gene_dtu_df %>% slice_head(n = 100) %>% pull(row_id)
top50_gene_dtu <- head(top100_gene_dtu, 50)
```

## Heatmap function for DTU -- note, it is not common to show DTU with heatmaps, so I am throwing in a volcano plot as well 
```{r plot-dtu-heatmap}
plot_dtu_heatmap <- function(mat, row_label_map, row_ids, title) {
  mat <- mat[rownames(mat) %in% row_ids, , drop = FALSE]
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  mat <- mat[apply(mat, 1, function(x) sd(x, na.rm = TRUE) > 0), , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  rownames(mat) <- row_label_map[rownames(mat)]

  annotation_df <- data.frame(
    sample_name = colnames(mat),
    group = ifelse(grepl("WT", colnames(mat)), "WT", "Q157R")
  ) %>% column_to_rownames("sample_name")

  annotation_colors <- list(group = c(Q157R = "#336B87", WT = "#E99787"))

  col_order <- c(grep("Q157R", colnames(mat), value = TRUE),
                 grep("WT", colnames(mat), value = TRUE))
  mat <- mat[, col_order]
  annotation_df <- annotation_df[col_order, , drop = FALSE]

  n_rows <- nrow(mat)
  plot_height <- min(10, 5 + n_rows * 0.15)
  fontsize_row <- if (n_rows <= 30) 10 else if (n_rows <= 60) 8 else 6

  pheatmap(mat, scale = "row", cluster_rows = TRUE, cluster_cols = FALSE,
           annotation_col = annotation_df, annotation_colors = annotation_colors,
           show_colnames = TRUE, show_rownames = TRUE, main = title,
           fontsize_row = fontsize_row, fontsize_col = 10, height = plot_height)
}
```

## Plot DTU Transcript Heatmaps
```{r plot-dtu-tx}
plot_dtu_heatmap(frac_matrix, frac_label_map, top100_tx_dtu, "Top 100 DTU Transcripts")
plot_dtu_heatmap(frac_matrix, frac_label_map, top50_tx_dtu, "Top 50 DTU Transcripts")
```

## Plot DTU Gene Heatmaps
```{r plot-dtu-genes}
plot_dtu_heatmap(gene_frac_matrix, gene_frac_label_map, top100_gene_dtu, "Top 100 DTU Genes")
plot_dtu_heatmap(gene_frac_matrix, gene_frac_label_map, top50_gene_dtu, "Top 50 DTU Genes")
```

## Interactive Volcano plot
```{r plot-dtu-volcano, echo=FALSE, message=FALSE}
# Clean and transform data
tx_dtu_df_clean <- tx_dtu_df %>%
  filter(!is.na(lf_DTU), !is.na(adj.p.value_DTU)) %>%
  mutate(
    # Extract PB accession and mark novel transcripts (same logic as frac_label_map)
    pb_from_transcript = str_extract(Transcript, "PB\\.[0-9]+\\.[0-9]+"),
    pb_from_index = str_extract(Index, "PB\\.[0-9]+\\.[0-9]+"),
    pb_accession = coalesce(pb_from_transcript, pb_from_index, Index),
    is_novel = !str_detect(Transcript, "ENSMUS"),
    transcript_label = ifelse(is_novel, paste0(pb_accession, "*"), pb_accession),
    # Calculate plotting variables
    log2_lf_DTU = sign(lf_DTU) * log2(abs(lf_DTU) + 1e-6),
    neg_log10_padj = -log10(adj.p.value_DTU),
    is_sig = adj.p.value_DTU <= 0.05,
    tooltip_label = paste0("Transcript: ", transcript_label,
                           "\nGene: ", Gene,
                           "\nlog2FC: ", signif(log2_lf_DTU, 3),
                           "\nFDR: ", signif(adj.p.value_DTU, 3))
  )

# Interactive volcano plot
gg_tx_dtu_interactive <- ggplot(tx_dtu_df_clean,
                                aes(x = log2_lf_DTU, y = neg_log10_padj)) +
  geom_point_interactive(
    aes(color = is_sig, tooltip = tooltip_label, data_id = row_id),
    size = 1.8, alpha = 0.7
  ) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Interactive DTU Volcano Plot (Q157R vs WT)",
    x = "Log2 Fold Change in Usage (Q157R vs WT)",
    y = "-Log10 Adjusted P-value",
    color = "adj.p ≤ 0.05"
  ) +
  theme(legend.position = "right")

# Render with ggiraph
girafe(ggobj = gg_tx_dtu_interactive, width_svg = 10, height_svg = 6)
```

#############################################################
########### Alternative Splice Event Heatmaps ##############
#############################################################
## Load Alternative Splicing Data
```{r load-as-data, include=FALSE}
as_df <- read_tsv("alternative_splice_summary.tsv") %>%
  mutate(
    row_id = make.unique(Index),
    adj_p = p.adjust(`Q157R-WT_p.value`, method = "BH"),
    is_sig = adj_p < 0.05,
    label = paste(Gene, Event_type, sep = " | ")
  )

psi_cols <- c("A258_Q157R_PSI", "X504_Q157R_PSI", "A309_Q157R_PSI",
              "V335_WT_PSI", "V334_WT_PSI", "A310_WT_PSI")

# PSI matrix for heatmap
as_matrix <- as_df %>%
  select(row_id, all_of(psi_cols)) %>%
  column_to_rownames("row_id") %>%
  as.matrix()

# Label map for rows: Gene | Event_type
as_label_map <- as_df %>%
  mutate(label = make.unique(label)) %>%
  select(row_id, label) %>%
  deframe()

# Significance map for row annotation
as_sig_map <- as_df %>%
  select(row_id, is_sig) %>%
  deframe()
```

## Define Alternative Splicing Subsets
```{r subset-as}
top_as_df <- as_df %>%
  arrange(`Q157R-WT_p.value`, desc(abs(`Q157R-WT_dPSI`)))

top100_as <- top_as_df %>% slice_head(n = 100) %>% pull(row_id)
top50_as <- head(top100_as, 50)
```
## Heatmap function for Alternative Splicing
```{r plot-as-heatmap}
plot_as_heatmap <- function(mat, row_label_map, row_ids, title) {
  # Filter matrix to include selected row_ids
  mat <- mat[rownames(mat) %in% row_ids, , drop = FALSE]
  
  # Keep rows with non-zero sum and some variability
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]
  mat <- mat[apply(mat, 1, function(x) sd(x, na.rm = TRUE) > 0), , drop = FALSE]
  if (nrow(mat) == 0) return(NULL)

  # Apply human-readable row labels (must be unique)
  rownames(mat) <- row_label_map[rownames(mat)]

  # Column annotation: WT vs Q157R
  annotation_df <- data.frame(
    sample_name = colnames(mat),
    group = ifelse(grepl("WT", colnames(mat)), "WT", "Q157R")
  )
  rownames(annotation_df) <- annotation_df$sample_name
  annotation_df$sample_name <- NULL

  # Annotation colors
  annotation_colors <- list(
    group = c(Q157R = "#336B87", WT = "#E99787")
  )

  # Column order: Q157R first, then WT
  col_order <- c(grep("Q157R", colnames(mat), value = TRUE),
                 grep("WT", colnames(mat), value = TRUE))
  mat <- mat[, col_order]
  annotation_df <- annotation_df[col_order, , drop = FALSE]

  # Dynamic plot sizing
  n_rows <- nrow(mat)
  plot_height <- min(10, 5 + n_rows * 0.15)
  fontsize_row <- if (n_rows <= 30) 10 else if (n_rows <= 60) 8 else 6

  # Final heatmap plot
  pheatmap::pheatmap(
    mat,
    scale = "row",
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    annotation_col = annotation_df,
    annotation_colors = annotation_colors,
    show_colnames = TRUE,
    show_rownames = TRUE,
    main = title,
    fontsize_row = fontsize_row,
    fontsize_col = 10,
    height = plot_height
  )
}
```

## Plot Alternative Splicing Heatmaps - these use a typical internal standard to show change
```{r plot-as-heatmaps}
plot_as_heatmap(as_matrix, as_label_map, top100_as, "Top 100 Splicing Events")
plot_as_heatmap(as_matrix, as_label_map, top50_as, "Top 50 Splicing Events")
```
## Alternative Splicing heatmaps with dPSI
```{r plot-as-dPSI-heatmap}
plot_psi_heatmap <- function(as_df, row_ids, title) {
  # Define PSI columns
  psi_cols <- c("A258_Q157R_PSI", "X504_Q157R_PSI", "A309_Q157R_PSI",
                "V335_WT_PSI", "V334_WT_PSI", "A310_WT_PSI")
  
  # Filter to selected row_ids
  filtered_df <- as_df %>%
    filter(row_id %in% row_ids) %>%
    select(row_id, label, all_of(psi_cols), is_sig)
  
  # Create PSI matrix
  psi_matrix <- filtered_df %>%
    select(row_id, all_of(psi_cols)) %>%
    column_to_rownames("row_id") %>%
    as.matrix()
  
  # Keep rows with non-zero sum and some variability
  psi_matrix <- psi_matrix[rowSums(psi_matrix, na.rm = TRUE) > 0, , drop = FALSE]
  psi_matrix <- psi_matrix[apply(psi_matrix, 1, function(x) sd(x, na.rm = TRUE) > 0), , drop = FALSE]
  
  if (nrow(psi_matrix) == 0) {
    warning("No valid PSI data found for the selected row_ids")
    return(NULL)
  }
  
  # Apply human-readable row labels (ensure uniqueness)
  remaining_row_ids <- rownames(psi_matrix)
  remaining_labels <- filtered_df$label[match(remaining_row_ids, filtered_df$row_id)]
  rownames(psi_matrix) <- make.unique(remaining_labels)
  
  # Clean up column names
  colnames(psi_matrix) <- gsub("_PSI", "", colnames(psi_matrix))
  
  # Column annotation: sample type
  col_annotation_df <- data.frame(
    group = ifelse(grepl("WT", colnames(psi_matrix)), "WT", "Q157R"),
    row.names = colnames(psi_matrix)
  )
  
  # Annotation colors
  annotation_colors <- list(
    group = c(Q157R = "#336B87", WT = "#E99787")
  )
  
  # Column order: Q157R first, then WT
  col_order <- c(grep("Q157R", colnames(psi_matrix), value = TRUE),
                 grep("WT", colnames(psi_matrix), value = TRUE))
  psi_matrix <- psi_matrix[, col_order]
  col_annotation_df <- col_annotation_df[col_order, , drop = FALSE]
  
  # Dynamic plot sizing
  n_rows <- nrow(psi_matrix)
  plot_height <- min(12, 6 + n_rows * 0.2)
  fontsize_row <- if (n_rows <= 30) 10 else if (n_rows <= 60) 8 else 6
  
  # Create heatmap
  pheatmap::pheatmap(
    psi_matrix,
    scale = "row",  # Scale by row to compare patterns across samples
    cluster_rows = TRUE,
    cluster_cols = FALSE,
    annotation_col = col_annotation_df,
    annotation_colors = annotation_colors,
    show_colnames = TRUE,
    show_rownames = TRUE,
    main = title,
    fontsize_row = fontsize_row,
    fontsize_col = 10,
    height = plot_height
  )
}
```
# Plot
```{r plot-as-dpsi-heatmaps}
plot_psi_heatmap(as_df, top50_as, "Top 50 Alternative Splicing Events")
plot_psi_heatmap(as_df, top100_as, "Top 100 Alternative Splicing Events")
```